\section{Related Work}

Our literature review focus on approaches similar to our main contribution, an interoperability layer for clinical research systems coupled with a front end.


\paragraph{SHRINE}

SHRINE (Shared Health Research Informatics NEtwork)~\cite{shrine} is a system that allows to network i2b2 instances together.
From a single webclient, a fork of the i2b2 webclient, several i2b2 can be queried at the same time, and their results be displayed in a unique interface.
By this, SHRINE offers an interoperability layer over an arbitrary number of i2b2 instances.
SHRINE's design is very tightly linked to i2b2 and extending its support to additional clinical research system is not possible.


\paragraph{Borderline}

% borderline
\emph{Borderline}~\cite{borderline} is a set of services that includes a user interface, and several back end components.
The user interface allows users to query clinical research platforms.
This UI communicates with its back end which manages the queries and proxies them to the connectors.
The connectors are responsible to query platforms containing patient data, such as tranSMART.
Borderline is a more of a query workflow management tool rather than a proper cohort explorer, and it is at the time of writing still a work-in-progress.



% check citations from other papers (bd2k and others)

% other systems:
% borderline:
% https://github.com/dsi-icl/borderline-server
% https://github.com/dsi-icl/borderline-middleware
% Etriks Analytics Environment
% https://eae.etriks.org/eae/


% - https://academic.oup.com/jamia/article/22/6/1132/2357622?searchresult=1
% - https://github.com/hms-dbmi/hackathon-Sept2017
% - http://dbmi.hms.harvard.edu/news/datathon-hackathon-sep-14-15
% - http://orthuber.com/MIE2009p0584.pdf
% - https://en.wikipedia.org/wiki/MultiDimensional_eXpressions
% - http://journals.sagepub.com/doi/abs/10.1177/1473871614526077
% - https://academic.oup.com/jamia/article/22/6/1114/2358127
% - https://www.nature.com/articles/sdata201696.pdf
% - https://www.sciencedirect.com/science/article/pii/S1532046417300709
% - 
% - CAVA: http://perer.org/papers/adamPerer-CAVA-IVS2014.pdf
% - http://orthuber.com/MIE2009p0584.pdf
% - explore refs of i2b2/shrine papers
% - is there transmart paper?
% - http://journals.sagepub.com/doi/abs/10.1177/1473871614526077: stopped at 18
% how to write: 
% - http://wiki.openi.org/openi-plug-in-for-i2b2

% todo: mention data sharing initiatives? all of us, eu initiatives


%they had a partnership with pic sure: %https://academic.oup.com/jamia/article/22/6/1132/2357622?searchresult=1
%pic sure 'wow' story (nhanes): %https://bd2kccc.org/wp-content/uploads/2017/02/15_PIC-SURE_FInal.pdf

